---
title: "SingleR Implementation for Cell Type Annotation"
author: "D. Ford Hannum"
date: "2023-07-12"
output: 
  html_document:
    toc: true
    highlight: tango
    number_sections: true
    toc_float: yes
    code_folding: hide
---
# Introduction

This file accompanies the Vizgen Analysis Guides: Cell Type Annotation.

Here we will look at the implementation of an open source package [SingleR](https://bioconductor.org/packages/release/bioc/html/SingleR.html) to assist in cell type annotation.

# Loading packages

```{r setup, include=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat) # to run normalizaiton methods
# BiocManager::install('SingleR')
# install.packages('matrixStats')
# BiocManager::install('scran')
library(scran)
library(SingleR)
library(data.table) # data processing library
library(ggplot2) # package to make visualizaitons
# BiocManager::install('scRNAseq')
library(scRNAseq) # package to access reference scRNA data
```


# Loading Data

Here we are going to use publicaly available data from a MERSCOPE mouse brain sample.

For our purposes we only need cell_by_gene.csv and cell_metadata.csv files.

```{r}
md <- fread('https://vzg-vzg-showcase-data.s3.amazonaws.com/Mouse_Brain_Showcase_232/Slice-1_Replicate-1/cell_metadata.csv',
               header = T, data.table = FALSE)

counts <- fread('https://vzg-vzg-showcase-data.s3.amazonaws.com/Mouse_Brain_Showcase_232/Slice-1_Replicate-1/cell_by_gene.csv',
               header = T, data.table = FALSE)

rownames(md) <- md$EntityID 
rownames(counts) <- counts$cell
counts <- counts[,-1] # removing cells as a column so we have a data frame of only counts
blanks <- counts[,grepl('Blank', colnames(counts))]
counts <- counts[,!grepl('Blank', colnames(counts))] # removing blanks

counts <- t(counts)

# counts[1:5,1:5]
# table(rownames(md) == colnames(counts)) # making sure the data frames align
```

```{r}
ggplot(md, aes(x = center_x, y = -center_y)) + 
  geom_point(size = .1) + theme_bw()
```

## Reference

Here we are going to use the [Chen et al. 2017](https://pubmed.ncbi.nlm.nih.gov/28355573/) mouse brain.

Below is a table of the cell type labels present in the data and the number of cells they have for each type.

```{r}
ref <- ChenBrainData()

table(ref$SVM_clusterID, exclude = F)
```

```{r}
# remotes::install_github("LTLA/scuttle")
library(scuttle)

ref <- logNormCounts(ref)
```


# Creating Seurat Object

We are going to create a Seurat object out of our data.

```{r, message = FALSE,}
spat <- CreateSeuratObject(counts = counts,
                           assay = 'RNA',
                           meta.data = md)
spat <- subset(spat, nCount_RNA >= 100)

spat <- NormalizeData(spat, normalization.method = 'LogNormalize',
                      scale.factor =  1e4)
spat <- ScaleData(spat, features = rownames(spat))

spat <- RunPCA(spat, features = rownames(spat), verbose = FALSE)

spat <- FindNeighbors(spat, dims = 1:10)

spat <- FindClusters(spat, resolution = .5)

spat <- RunUMAP(spat, dims = 1:10)

```

# Running SingleR

```{r}
# We need the Seurat object as a single cell expeirment for SingleR
test <- as.SingleCellExperiment(spat)

# Running SingleR
pred <- SingleR(test = test, ref = ref, labels = ref$SVM_clusterID,
                de.method = 'wilcox')

spat$labels_fine <- pred$labels
spat$delta <- pred$delta.next
spat$labels_prune <- pred$pruned.labels
```

Below is a table of the predicted labels.

```{r}
table(pred$labels)
```

## General Groups

Here I'm going to combine some of the groups into larger groups for easy of analysis.

The larger groups I'm going to create are GABA, Glu, and Epith; which combine the many GABA+, Glu+, and Epith labels.

```{r}
spat$labels_general <- ifelse(grepl('GABA', spat$labels_fine),'GABA',
                              ifelse(grepl('Glu', spat$labels_fine),'Glu',
                                     ifelse(grepl('Epith', spat$labels_fine), 'Epith',spat$labels_fine)))

# table(spat$labels_fine, spat$labels_general)
```


# Viewing labels

We can view how the labels look by comparing them to clusters, viewing in UMAP coordinates, and viewing with the spatial context.

## Clusters {.tabset}

We would expect cell type labels to group strongly with clusters.


```{r}
df <- as.data.frame(table(spat$seurat_clusters, spat$labels_general))
# df

df$Var1.sum <- NA

for (i in unique(df$Var1)){
  df[df$Var1 == i,]$Var1.sum <- sum(df[df$Var1 == i,]$Freq)
}

df$Var2.sum <- NA

for (i in unique(df$Var2)){
  df[df$Var2 == i,]$Var2.sum <- sum(df[df$Var2 == i,]$Freq)
}

df$Var1.perc <- df$Freq / df$Var1.sum
df$Var2.perc <- df$Freq / df$Var2.sum

df$Jaccard <- df$Freq / (df$Var1.sum + df$Var2.sum - df$Freq)

order_ <- c()
cnts_ <- c()
for (i in unique(df$Var1)){
  temp <- df[df$Var1 == i,]
  order_ <- c(order_,which.max(temp$Freq))
  cnts_ <- c(cnts_,max(temp$Freq))
}
temp <- data.frame(row.names = unique(df$Var1),
                   order = order_,
                   cnts = cnts_)
df$Var1_ordered <- factor(df$Var1,
                          levels = rownames(temp[order(temp$order, -temp$cnts),]))

df$`Cluster\nPercentage` <- df$Var1.perc

ggplot(df, aes(x = Var1_ordered, y = Var2, label = Freq, fill = `Cluster\nPercentage`)) +
  geom_tile() +
  xlab('Clusters') + ylab('Cell Labels') +
  geom_text(size = 3) +
  scale_fill_gradient2(low = 'white', mid = "pink", high = 'red',
                       midpoint = .5) +
  ggtitle('Cross-Tabulation of Clusters vs Predicted Cell Labels')
```

We see that some clusters are made up of primarily one type of cell (ex. cluster 1 is almost exclusively astro), whereas others show a combination of multipe cell types (ex. cluster 16 is made up in large parts by epith, GABA, and Glu).

## UMAP

Below is the general UMAP plot with the labels added.

```{r}
DimPlot(spat)

DimPlot(spat, group.by = 'labels_general', label = T)
```

### Viewing one cell type at a time

```{r}
cols_ <- scales::hue_pal()(length(unique(spat$labels_general)))

names(cols_) <- levels(as.factor(spat$labels_general))
```

```{r, fig.show='hold', out.width='50%'}
for (i in levels(as.factor(spat$labels_general))){
  # print(i)
  temp_cols <- ifelse(names(cols_) == i,cols_[i], 'lightgrey')
  counts_ <- sum(spat$labels_general == i)
  print(DimPlot(spat, group.by = 'labels_general') +
          scale_color_manual(values = temp_cols) +
          ggtitle(paste0(i, ' (n = ',counts_,')')))
}
```

## Spatial

Now we are going to look at the labels for the cells with their spatial coordinates

```{r}
ggplot(spat@meta.data, aes(x= center_x, y = -center_y)) +
  geom_point(size = .1) +
  theme_bw() + xlab('') + ylab('')
```

```{r}
ggplot(spat@meta.data, aes(x= center_x, y = -center_y, colour = labels_general)) +
  geom_point(size = .1) +
  theme_bw() + xlab('') + ylab('') +
  guides(colour = guide_legend(override.aes= list(size = 4)))
```

### Viewing one cell type at a time

```{r, fig.show='hold', out.width='50%'}
temp <- spat@meta.data
for (i in levels(as.factor(spat$labels_general))){
  # print(i)
  temp_cols <- ifelse(names(cols_) == i,cols_[i], 'lightgrey')
  counts_ <- sum(spat$labels_general == i)
  temp$order <- temp$labels_general == i
  temp <- temp[order(temp$order),]
  temp$size <- ifelse(temp$order, 1,.1)
  print(ggplot(temp, aes(x= center_x, y = -center_y, colour = labels_general,
                         size = size)) +
          geom_point() +
          scale_size_identity() +
          theme_bw() + xlab('') + ylab('') +
          scale_color_manual(values = temp_cols) +
          guides(colour = guide_legend(override.aes= list(size = 4))))
}

```

